% Progress report for S. Banerjee - ARC-AGI-3 world model + RL
% Compile: pdflatex progress-report-soumya.tex  (run twice for refs)
\documentclass[10pt]{article}
\usepackage[a4paper,margin=1.9cm]{geometry}
\usepackage{booktabs}
\usepackage{amsmath}
\usepackage{microtype}
\usepackage[hidelinks]{hyperref}
\usepackage{titlesec}
\titlespacing*{\section}{0pt}{1.0ex}{0.5ex}
\titlespacing*{\subsection}{0pt}{0.7ex}{0.3ex}
\setlength{\parskip}{0.25em}
\setlength{\parindent}{0pt}
\pagestyle{empty}

\title{\vspace{-1.6cm}A Gymnasium Substrate and World-Model Baseline for ARC-AGI-3:\\
\large DreamerV3 Fits the World but Cannot Yet Act in It}
\author{Hasaan Ahmad \quad\textbar\quad Supervisor: Soumya Banerjee (University of Cambridge)}
\date{Progress report --- \today}

\begin{document}
\maketitle
\vspace{-1.2em}

\section{Motivation and gap}
ARC-AGI-3 is a benchmark of 25 public interactive abstract-reasoning games in
which the rules and goals are \emph{not} given and must be inferred through
active play. It is scored by RHAE (Relative Human Action Efficiency): per level,
$s_i=\min\big((\text{human\_actions}/\text{ai\_actions})^2,\,1.15\big)$; a game
score is the level-index-weighted mean $\sum_i s_i\, i / \sum_i i$ over all
levels (including uncompleted ones), and the total score averages games equally.
The ARC-AGI living survey~\cite{survey2026} ($\sim$80 papers across
ARC-AGI-1/2/3) documents a consistent $\sim$3$\times$ performance cliff into
ARC-AGI-3, finds only 3 of 80 papers report any ARC-AGI-3 result --- \emph{none}
world-model based --- and names world-model induction as the required next step.
Ying et al.~\cite{ying2025} argue the same from first principles: current
evaluations test static pre-trained representations, whereas the capability of
interest is using an abstract world model as a prior to build task-specific ones
on the fly, for which ARC-AGI-3-style games are the proposed testbed. Lee et
al.~\cite{lee2024} give the one direct precedent --- DreamerV3 beating PPO on
multi-step ARC-AGI-1 --- and call for scaling to harder interactive tasks. This
work is the first model-based RL (MBRL) entry on ARC-AGI-3 and supplies the
missing standard RL entry point.

\section{Contributions}
\textbf{(1) Substrate (primary artifact).} The first Gymnasium- and
DreamerV3-\texttt{embodied}-compatible wrapper for ARC-AGI-3: a 4102-way flat
discrete action space (parameter-less \texttt{ACTION1--5}; \texttt{ACTION6}
over the $64\!\times\!64$ click grid; \texttt{ACTION7}/undo) with per-game
boolean masking on the actor logits; a $(64,64,3)$ \texttt{uint8} image
observation in DreamerV3's native modality; an offline loader for the
340-replay human corpus; and a post-hoc RHAE harness. Action coding, loader,
and RHAE are property-/regression-tested (318-test suite green). DreamerV3 is
used \emph{stock and unforked} (\texttt{size12m}, the ARC-AGI-1-precedented
choice~\cite{lee2024}), registered as an \texttt{embodied} environment with no
upstream modification.

\textbf{(2) Controlled negative result.} A paired baseline of stock DreamerV3
on 6 public games $\times$ 2 seeds, from-scratch vs.\ warm-started from a
cross-game world model pretrained on all 340 replays (Regime B: shared WM,
per-game actor). Cross-game pretraining yields no measurable benefit within
seed variance.

\textbf{(3) Mechanistic diagnosis.} World-model reconstruction and dynamics
losses collapse to convergence while RHAE stays $\approx$0: the model learns
to predict the environment, but the controller cannot exploit it in
imagination within the budget.

\section{Method}
The world model (encoder + RSSM + decoders + reward/continue heads) was
pretrained self-supervised on the full mixed 340-replay buffer (1$\times$A100;
image-reconstruction loss $53.66\!\to\!0.14$). Per game, the pretrained WM
initialises a run with a fresh actor and critic, the game's human replays
pre-populate the buffer, and online interaction proceeds for 500k env-steps at
default DreamerV3 hyperparameters; the cold arm is identical but for removing
the warm-start, making it the \emph{only} difference between arms --- a clean
ablation of the Regime-B premise. RHAE human baselines use the upper-median of
first-time-player action counts, dropping levels with $<2$ completers from both
numerator and denominator (coverage 70.5\%, 129/183 levels).

\section{Results}

\begin{table}[h]
\centering
\small
\caption{Paired RHAE @ 500{,}000 env-steps. Stock DreamerV3 \texttt{size12m},
6 public games $\times$ 2 seeds, cold (from-scratch) vs.\ warm (cross-game
WM-pretrained). $\Delta=$ warm$-$cold. Higher is better; human-parity $\approx$
the per-level $1.15$ cap.}
\label{tab:rhae}
\begin{tabular}{lcccccc}
\toprule
& \multicolumn{2}{c}{Cold} & \multicolumn{2}{c}{Warm} & & \\
\cmidrule(lr){2-3}\cmidrule(lr){4-5}
Game & s0 & s1 & s0 & s1 & Cold $\bar{x}$ & Warm $\bar{x}$ \\
\midrule
vc33 & 0.0411 & 0.0182 & 0.0548 & 0.0166 & 0.0297 & 0.0357 \\
sb26 & 0.0000 & 0.0000 & 0.0000 & 0.0000 & 0.0000 & 0.0000 \\
cd82 & 0.0000 & 0.0000 & 0.0000 & 0.0000 & 0.0000 & 0.0000 \\
tn36 & 0.0000 & 0.0000 & 0.0000 & 0.0000 & 0.0000 & 0.0000 \\
ls20 & 0.0000 & 0.0000 & 0.0000 & 0.0000 & 0.0000 & 0.0000 \\
lf52 & 0.0000 & 0.0000 & 0.0000 & 0.0000 & 0.0000 & 0.0000 \\
\midrule
\textbf{Combined (mean of 6)} & & & & & \textbf{0.0049} & \textbf{0.0059} \\
\bottomrule
\end{tabular}
\end{table}

\vspace{-0.4em}
Only vc33 is ever non-zero, and there the per-seed sign of $\Delta$ disagrees
(s0 favours warm by $0.014$, s1 favours cold by $0.002$): at $n=2$ the
warm$-$cold delta lies within seed variance, so the Regime-B premise that
cross-game pretraining aids downstream control is \emph{not supported}. The
pre-registered gate (RHAE $>0$ on $\ge 2/3$ of \{vc33,\,sb26,\,cd82\}) failed
$1/3$ --- a clean, paired, 6-game negative result, not one unlucky run.

\begin{table}[h]
\centering
\small
\caption{The dissociation: the world model fits \emph{better} on a game the
controller never solves (cd82) than on the only game it ever solves (vc33).
Both columns are seed-0 warm runs from the same paired experiment as
Table~\ref{tab:rhae}; the pattern holds across all 6 pilot warm cells.}
\label{tab:diag}
\begin{tabular}{lcc}
\toprule
Signal & vc33 (only solved) & cd82 (never solved) \\
\midrule
WM image-recon.\ loss & $58.3 \to 0.16$ ($-99.7\%$) & $31.0 \to \mathbf{0.06}$ ($-99.8\%$) \\
WM dynamics loss & $10.6 \to 1.24$ & $11.4 \to 1.03$ \\
Train eps.\ clearing $\ge$1 lvl & $17/9751$ & $0/4950$ \\
Eval eps.\ clearing $\ge$1 lvl & $4/23$ & $0/24$ \\
Random-action frac.\ @\,500k & $0.750$ & $\mathbf{1.000}$ \\
Post-hoc RHAE & $0.0548$ & $0.0000$ \\
\bottomrule
\end{tabular}
\end{table}

\vspace{-0.4em}
Table~\ref{tab:diag} is the central finding in one view. cd82's world model
fits \emph{tighter} than vc33's (image loss $0.06$ vs.\ $0.16$) --- yet vc33 is
the only game the controller ever solves, while cd82's policy never leaves a
$1.000$ random-action fraction across the entire 500k-step run. The model
learns the environment; the controller cannot exploit it. This holds across all
six pilot warm cells (\{vc33, sb26, cd82\}$\times$2 seeds): image loss collapses
$>$95\% everywhere, while both gate-failure games clear $0/24$ eval episodes in
every cell. The bottleneck is the controller, not the world model --- not a
single lucky or unlucky run.

\section{Discussion and positioning}
This is the first MBRL data point on ARC-AGI-3 and sharpens the survey's
$\sim$3$\times$ cliff~\cite{survey2026}: a model that provably fits these
environments still cannot act in them under a stock controller at a realistic
budget --- precisely the ``static representation vs.\ on-the-fly task model''
gap of Ying et al.~\cite{ying2025}. It also qualifies the Lee et
al.~\cite{lee2024} precedent: DreamerV3's ARC-AGI-1 success does \emph{not}
transfer to the interactive, rule-inference regime at this scale, and
cross-game pretraining does not close the gap --- isolating the controller, not
the world model, as the locus of difficulty. \textbf{Limitations:} a compute-bounded 500k-step budget and $N=6\times2$ ---
a scoped claim about \emph{stock DreamerV3 at this budget $\pm$ this
pretraining}, not that MBRL cannot do ARC-AGI-3; the diagnosis carries the
weight, not the headline number.
\textbf{Status and planned work.} The substrate (contribution~1) is complete,
tested, and reusable independent of any result; the controlled negative result
and its diagnosis (2--3) are established across the full 6-game paired sweep.
Three steps remain, each with a pre-specified criterion that would confirm or
relocate the diagnosis. \emph{(i)~Representational probing:} linear decoders
trained on the frozen RSSM latents for task-relevant variables --- level
identity, level-transition events, and object configuration --- against a
label-permuted control; above-control accuracy confirms the world model
encodes the structure the controller fails to exploit, whereas a null result
would relocate the bottleneck from control to representation learning.
\emph{(ii)~Generative-rollout fidelity:} Fr\'echet Video Distance between
imagined and ground-truth trajectories over the actor's planning horizon,
establishing whether the model's predictions remain faithful across the
rollout length on which the policy is trained. \emph{(iii)~Controller-side
intervention:} a bounded ablation that raises the imagination-to-environment
training ratio and steepens the entropy-annealing schedule with the world
model held fixed, isolating whether the gap closes under additional
policy-improvement compute alone rather than a change of model class. Steps
(i)--(ii) are compute-light and operate on a frozen model; (iii) is a single
bounded retraining sweep. The substrate contribution is independent of all
three outcomes.

\begin{thebibliography}{9}
\small
\setlength{\itemsep}{0pt}
\bibitem{lee2024} D.\ Lee et al.\ \emph{Reinforcement Learning and Model-Based
Reasoning on the Abstraction and Reasoning Corpus.} arXiv:2408.14855, 2024.
\bibitem{survey2026} \emph{The Abstraction and Reasoning Corpus: A Living
Survey.} arXiv:2603.13372, 2026.
\bibitem{ying2025} L.\ Ying et al.\ \emph{Instance vs.\ Abstract World Models.}
arXiv:2507.12821, 2025.
\bibitem{dreamerv3} D.\ Hafner, J.\ Pasukonis, J.\ Ba, T.\ Lillicrap.
\emph{Mastering Diverse Control Tasks through World Models.} Nature, 2025.
arXiv:2301.04104.
\bibitem{arc3} \emph{ARC-AGI-3: Interactive Reasoning Benchmark.}
arXiv:2603.24621, 2026.
\end{thebibliography}

\end{document}
